---
title: "Statistical power of multivariate GWAS"
author: "by [Baptiste Couvy-Duchesne] - `r format(Sys.time(), '%d %B %Y')`"
output:
  epuRate::PCTG:
    toc: TRUE
    code_folding: "show"
---

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# You need these libraries to run this template:
library(rmarkdown)    # install.packages("rmarkdown") 
library(epuRate)      # devtools::install_github("holtzy/epuRate", force=TRUE)

# Multivariate power requires
library(MBESS)
library(Matrix)
library(matrixcalc)
library(mvtnorm)
library(SCGLR)
```

<br><br><br><br>

> Code associated with the manuscript "A fast method for estimating statistical power of multivariate GWAS in real case scenarios: examples from the field of imaging genetics"  

# Functions used to calculate power 

## Univariate power

```{r, message=FALSE}

calculateUnivPowerAnalytic<-function(R2, MAF, alpha, Ns){

  # Initialise  
pwF<-0
wF<-0
FF<-0

for (N in Ns){

sdY<-1 # Variance phenotype assumed to be 1
r<-1 
a<-1
b<-1
C<-diag(1, nrow = a, ncol=r)

# Critical value from an inverse central distribution F:
Fcrit<-qf(1-alpha, a, N-r)
FF<-c(FF, Fcrit)
# Non centrality parameter
w<-Rsquare2Lambda(R2 = R2, N = N)
wF<-c(wF, w)
# Power
pw<-1-pf(Fcrit, a, N-r, w)
pwF<-c(pwF, pw)

}
return(pwF)
}


```

## Calculate Sigma

```{r, message=FALSE}

calculateSigma<-function(R2, aPaths,MAF, ePaths, factorNb, nvar){

AA=aPaths %*% t(aPaths) # CovA
EE=ePaths %*% t(ePaths) # CovE
RR=AA+EE # CovP

# Effect of latent trait on phenotypes
betaA2<-sqrt(R2)/max(abs(aPaths[,factorNb]))* sign(aPaths[,factorNb][which( abs(aPaths[,factorNb]) == max(abs(aPaths[,factorNb]))) ] )
betaVect<-betaA2 * aPaths[,factorNb]

# Variance and covariance due to the effect removed from A
AA2= AA - betaVect %*% t(betaVect)

# Residual variance-covariance matrix
RR2= AA2 + EE 
return(RR2)
}

```

## MANOVA power calculation 

```{r, message=FALSE}
estimatePowerMultivariateConservative<-function(R2, MAF, aPaths, ePaths, nvar, Ns, alpha, method){

   start_time=Sys.time()
pwPBfinal<-NULL
pwWfinal<-NULL
pwHLTfinal<-NULL
r<-1
a<-1
b<-nvar
s<-min(a,b)
C<-diag(1, nrow = a, ncol=r)
sdY<-1

for (factorNb in 1:nvar){
  
sigma<-calculateSigma(R2 = R2,MAF=MAF, aPaths= aPaths, ePaths=ePaths, factorNb = factorNb, nvar = nvar)

pwPBf<-NULL
pwWf<-NULL
pwHLTf<-NULL
sdSNP<-1

for (N in Ns){

# Fcrit and df of residuals
df2PB<-s*(N-r-b+s)
FcritPB<-qf(1-alpha, df1=a*b, df2=df2PB)

g<-sqrt(((a**2)*(b**2)-4)/(a**2 + b**2-5))
df2W<-g*((N-r) - (b-a+1)/2) - (a*b-2)/2
FcritW<-qf(1-alpha, df1=a*b, df2=df2W)

df2HLT<-s*((N-r) -b -1 )+2
FcritHLT<-qf(1-alpha, df1=a*b, df2=df2HLT)

SNP <-rbinom(N, 2, MAF)
SNP= (SNP - mean(SNP)) / sd(SNP)
SNP<-matrix(SNP, nrow=N, ncol=1)

Cmat<-matrix(sdSNP**2, nrow=1, ncol=1)
Umat<-diag(1, nrow=nvar, ncol=nvar)

betaA<-sqrt(R2)*sdY/sdSNP/max(abs(aPaths[,factorNb])) * sign(aPaths[,factorNb][which( abs(aPaths[,factorNb]) == max(abs(aPaths[,factorNb]))) ] )
betaVect<-matrix(betaA * aPaths[,factorNb], nrow=1, ncol=nvar)

thetaH1<- Cmat %*% betaVect %*% Umat

Hmat<-t(thetaH1) %*% solve(Cmat %*% solve(t(SNP) %*% SNP) %*% t(Cmat)) %*% thetaH1
Emat<- (t(Umat) %*% sigma %*% Umat) * (N-r)
Tmat<-Hmat + Emat

# Trace Pillai-Bartlett
PBstat<- matrix.trace(Hmat %*% solve(Tmat))
ncpPB<-(PBstat/s)/((1-PBstat/s)/df2PB)
pwPB<-1-pf(FcritPB, a*b, df2PB, ncpPB)
pwPBf<-cbind(pwPBf, pwPB)

# Wilks' lambda
Wstat<-det(Emat %*% solve(Tmat))
ncpW<-(1-Wstat**(1/g))/(Wstat**(1/g) / df2W)
pwW<-1-pf(FcritW, a*b, df2W, ncpW)
pwWf<-cbind(pwWf, pwW)

# Hotelling-Lawley trace
HLTstat<-matrix.trace(Hmat %*% solve(Emat))
ncpHLT<-(HLTstat/s)/(1/df2HLT)
pwHLT<-1-pf(FcritHLT, a*b, df2HLT, ncpHLT)
pwHLTf<-cbind(pwHLTf, pwHLT)

}

pwPBfinal<-rbind(pwPBfinal, pwPBf)
pwWfinal<-rbind(pwWfinal, pwWf)
pwHLTfinal<-rbind(pwHLTfinal, pwHLTf)

}
result<-rbind(c(0,colMeans(pwPBfinal, na.rm = T)), c(0,colMeans(pwWfinal, na.rm = T)), c(0,colMeans(pwHLTfinal, na.rm = T)))

# Get A estimates to weight power by factors
# apaths are standardised, so A and E are pretty straighforward to calculate
wA<-colSums(aPaths * aPaths)
#print(wA)

# Print the number of genetic factors that cannot handle a specific effect size (they explain less than R2 on any of the variables)
print(paste(length(which(is.na(pwPBfinal[,1]))), "Factors were excluded for R2=", R2 ))

# Exclude these factors
if (length(which(is.na(pwPBfinal[,1])))>0){
wA<-wA[-which(is.na(pwPBfinal[,1]))]
pwPBfinal<-pwPBfinal[-which(is.na(pwPBfinal[,1])),]
pwWfinal<-pwWfinal[-which(is.na(pwWfinal[,1])),]
pwHLTfinal<-pwHLTfinal[-which(is.na(pwHLTfinal[,1])),]
}
wA<-wA/sum(wA)
resultWA<-rbind(c(0,wA %*% pwPBfinal), c(0, wA %*% pwWfinal), c(0,wA %*% pwHLTfinal))

rownames(resultWA)<-c("PB", "W", "HLT")
colnames(resultWA)<-c(0,Ns)

 end_time=Sys.time()
 print(paste0("Total time spent computing power: ", end_time - start_time))
 
 if (method=="factorWise"){ return(signif(pwPBfinal, 3) ) }
  if (method=="weights"){ return(signif(resultWA, 3) ) }
 if (method=="noweights"){return(signif(rbind(c(0,colMeans(pwPBfinal)), c(0,colMeans(pwWfinal)), c(0,colMeans(pwHLTfinal)) ),3))}

}

```


## Simulate power 

```{r, message=FALSE}

 powerSimulatedMANOVA=function(Ns, MAF, aPaths, ePaths, nvar, R2, alpha){   
  
  empPwrAllN=NULL
  sdY<-1
  start_time=Sys.time()

  for (N in Ns){
  empPwr=NULL
  print(paste0("N= ", N ) )
   
  for (factorNb in 1:nvar){
    
    print(paste0("Factor: " , factorNb))
    pall=NULL
    
for (simul in 1:100){  
  # Simulate SNP  
SNP <-rbinom(N, 2, MAF)
SNP<-matrix(SNP, nrow=N, ncol=1)
sdSNP<-sqrt(2*MAF*(1-MAF))

# Vector of betas and corresponding A and E
betaA<-sqrt(R2)*sdY/sdSNP/max(abs(aPaths[,factorNb]))*sign(aPaths[,factorNb][which( abs(aPaths[,factorNb]) == max(abs(aPaths[,factorNb]))) ] )
betaVect<-matrix(betaA * aPaths[,factorNb], nrow=1, ncol=nvar)
Ax<-betaA*SNP + rnorm(n=N, mean=0, sd=sqrt(1-R2)  ) 

A<-rmvnorm(n=N, mean=rep(0,nvar) , sigma=diag(1, nvar, nvar))
E<-rmvnorm(n=N, mean=rep(0,nvar) , sigma=diag(1, nvar, nvar))
A[,factorNb]=Ax

# Phenotypes
P<- A %*% t(aPaths) + E %*% t(ePaths)

# Perform test
m0=manova(P ~ SNP )
pv=summary(m0, test="Pillai")$stats[1,6]
pall=c(pall, pv)

}
  empPwr=rbind(empPwr,length(which(pall<alpha)) / length(pall) )
   }
    empPwrAllN=cbind(empPwrAllN, empPwr)
  }
  
end_time=Sys.time()
print(paste0("Total time spend computing power: ", end_time - start_time, " seconds"))   
return(empPwrAllN)
  
}
```

## Wrap up functions to compute all the power curves

```{r, message=FALSE}

# Analytical power
calculateAndPlotPower<-function(aPathsFile, ePathsFile, codeName, nvarIndep, nvar, alpha, method){

aPaths<-read.csv(aPathsFile)
aPaths$X<-NULL
aPaths<-as.matrix(aPaths)
ePaths<-read.csv(ePathsFile)
ePaths$X<-NULL
ePaths<-as.matrix(ePaths)

MAF<-0.02
alpha<-5e-8

Ns<-c(1000,2000,3000,4000,5000,6000,7000,8000,9000,10000,15000,20000,25000,30000,35000,40000,45000,50000,55000, 60000)
R2s<-c(0.01,0.0075,0.005,0.0025,0.001,0.00075,0.0005)
for (R2 in R2s){
write.csv(estimatePowerMultivariateConservative(R2 = R2, MAF = MAF, aPaths = aPaths, ePaths = ePaths, nvar = nvar, Ns = Ns, alpha = alpha, method=method), file = paste0("ACE_may2018/PowerAnalytic_multiv_" , codeName,"_", R2, "_", method, ".csv"))
}
}

# Simulated power
calculateAndPlotPowerSimulated<-function(aPathsFile, ePathsFile, codeName, nvar, alpha){

aPaths<-read.csv(aPathsFile)
aPaths$X<-NULL
aPaths<-as.matrix(aPaths)
ePaths<-read.csv(ePathsFile)
ePaths$X<-NULL
ePaths<-as.matrix(ePaths)

MAF<-0.02
Ns<-c(1000,2000,3000,4000,5000,6000,7000,8000,9000,10000,15000,20000,25000,30000,35000,40000,45000,50000,55000, 60000)
R2s<-c(0.01,0.0075,0.005,0.0025,0.001,0.00075,0.0005)

for (R2 in R2s){
write.csv(powerSimulatedMANOVA(R2 = R2, MAF = MAF, aPaths = aPaths, ePaths = ePaths, nvar = nvar, Ns = Ns, alpha = alpha), file = paste0("ACE_may2018/PowerAnalytic_multiv_" , codeName,"_", R2, "_Simulated.csv"))
}
}

```


# Run analyses

```{r, message=FALSE}

calculateAndPlotPower(aPathsFile = "ACE_dec2016/a_sign_meanLR_SubVols_ICVreg.csv", ePathsFile = "ACE_dec2016/e_sign_meanLR_SubVols_ICVreg.csv", codeName = "MeanLR_subcortVolumes_plotFucntion", nvarIndep = 6, nvar = 7, alpha=5e-8, method="weights")

#calculateAndPlotPower(aPathsFile = "ACE_dec2016/a_sign_meanLR_hippoSF_ICVreg.csv", ePathsFile = "ACE_dec2016/e_sign_meanLR_hippoSF_ICVreg.csv", codeName = "MeanLR_hippoSubfield_plotFucntion", nvarIndep = 7, nvar = 12, alpha=5e-8, method="weights")

# Simulation results
#calculateAndPlotPowerSimulated(aPathsFile = "ACE_dec2016/a_sign_meanLR_SubVols_ICVreg.csv", ePathsFile = "ACE_dec2016/e_sign_meanLR_SubVols_ICVreg.csv", codeName = "MeanLR_subcortVolumes_plotFucntion", nvar = 7, alpha=5e-8)

#calculateAndPlotPowerSimulated(aPathsFile = "ACE_dec2016/a_sign_meanLR_hippoSF_ICVreg.csv", ePathsFile = "ACE_dec2016/e_sign_meanLR_hippoSF_ICVreg.csv", codeName = "MeanLR_hippoSubfield_plotFucntion",  nvar = 12, alpha=5e-8)

```

# Example of plot

## Plot subcortical volumes

```{r, message=FALSE}

# Plot
pw1<-read.csv("ACE_may2018/PowerAnalytic_multiv_MeanLR_subcortVolumes_plotFucntion_0.01_weights.csv", header=T, row.names = 1)
pw2<-read.csv("ACE_may2018/PowerAnalytic_multiv_MeanLR_subcortVolumes_plotFucntion_0.0075_weights.csv", header=T, row.names = 1)
pw3<-read.csv("ACE_may2018/PowerAnalytic_multiv_MeanLR_subcortVolumes_plotFucntion_0.005_weights.csv", header=T, row.names = 1)
pw4<-read.csv("ACE_may2018/PowerAnalytic_multiv_MeanLR_subcortVolumes_plotFucntion_0.0025_weights.csv", header=T, row.names = 1)
pw5<-read.csv("ACE_may2018/PowerAnalytic_multiv_MeanLR_subcortVolumes_plotFucntion_0.001_weights.csv", header=T, row.names = 1)
pw6<-read.csv("ACE_may2018/PowerAnalytic_multiv_MeanLR_subcortVolumes_plotFucntion_0.00075_weights.csv", header=T, row.names = 1)
pw7<-read.csv("ACE_may2018/PowerAnalytic_multiv_MeanLR_subcortVolumes_plotFucntion_5e-04_weights.csv", header=T, row.names = 1)

Ns<-c(1000,2000,3000,4000,5000,6000,7000,8000,9000,10000,15000,20000,25000,30000,35000,40000,45000,50000,55000, 60000)
R2s<-c(0.01,0.0075,0.005,0.0025,0.001,0.00075,0.0005)
MAF=0.2
nvar=7

# PLOT
#png("ACE_may2018/Power_UnivVSMultiv_analytic_MLR_subCortVol_ICVreg_conservative_weight.png",  width = 28, height = 15, units = "cm", res=400)
par(xpd=F, mar=par()$mar+c(4,0,-3,0))
plot(c(0,Ns)/1000, pw1[1,], pch=20, xlab="Sample size (in thousands)", ylab="Power", ylim=c(0,1))
library(scales)
NNs<-7

colM<-hue_pal()(NNs)
for (iii in 1:NNs){
  points(c(0,Ns)/1000,get(paste("pw", iii, sep = ""))[2,], col=colM[iii], pch=20)
  lines(c(0,Ns)/1000,get(paste("pw", iii, sep = ""))[2,], col=colM[iii], lwd=2)

   points(c(0,Ns)/1000,calculateUnivPowerAnalytic(R2 = R2s[iii], MAF = MAF, alpha = 5e-8/6, Ns = Ns), col=colM[iii], pch=2)
  lines(c(0,Ns)/1000,calculateUnivPowerAnalytic(R2 = R2s[iii], MAF = MAF, alpha = 5e-8/6, Ns = Ns), col=colM[iii], lty=2, lwd=2)

     points(c(0,Ns)/1000,calculateUnivPowerAnalytic(R2 = R2s[iii], MAF = MAF, alpha = 5e-12, Ns = Ns), col=colM[iii], pch=8)
  lines(c(0,Ns)/1000,calculateUnivPowerAnalytic(R2 = R2s[iii], MAF = MAF, alpha = 5e-12, Ns = Ns), col=colM[iii], lty=3, lwd=2)

}
grid()
abline(h=0.8, lty=2, lwd=2)
par(xpd=T)
legend(x = -3000/1000, y = -0.3, legend = c("R2=1%","R2=0.75%", "R2=0.5%", "R2=0.25%", "R2=0.1%", "R2=0.075%", "R2=0.05%"), col=colM, pch=19, cex=1, horiz = T, box.col = "white")
legend(x = -3000/1000, y = -0.4, legend = c("Multivariate","Univariate 5e-8/neff   ", "Univariate 5e-12"), col="black", lty=c(1,2,3), cex=1,pch=c(20, 2,8), horiz = T, box.col = "white")
#dev.off()

```

# Sensitivity analysis

```{r, message=FALSE}

aPathsFile = "QTIM_results/sta_matrix_subcortical.csv"
ePathsFile = "QTIM_results/ste_matrix_subcortical.csv"
aPaths<-read.csv(aPathsFile)
aPaths$X<-NULL
aPaths<-as.matrix(aPaths)
ePaths<-read.csv(ePathsFile)
ePaths$X<-NULL
ePaths<-as.matrix(ePaths)

aPathsSE=read.csv("QTIM_results/sta_SE_matrix_subcortical.csv")
aPathsSE$X<-NULL
aPathsSE<-as.matrix(aPathsSE)

ePathsSE=read.csv("QTIM_results/ste_SE_matrix_subcortical.csv")
ePathsSE$X<-NULL
ePathsSE<-as.matrix(ePathsSE)


# Initiate plot 

#tiff(filename = "SensitivityAnalysis_Weights_subcorticalVolumes.tiff",compression = "lzw+p", width = 28, height = 15, units = "cm", res=400)

nvar=7
Ns = c(1000,2000,3000,4000,5000,6000,7000,8000,9000,10000,15000,20000,25000,30000,35000,40000,45000,50000,55000, 60000)
pw0=estimatePowerMultivariateConservative(R2 = 0.01, MAF = 0.2, aPaths = aPaths, ePaths = ePaths, nvar = 7, Ns = Ns, alpha = 5e-8, method = "weights")

# Initialise plot
par(xpd=F, mar=par()$mar+c(3,0,-3,0))
plot(c(0,Ns), pw0[1,], pch=20,  xlab="Sample size (in thousands)", ylab="Power", ylim=c(0,1))
lines(c(0,Ns), pw0[1,])
colM=hue_pal()(7)

# Loop over all effect sizes
for (iii in 1:7){

# Vary aPaths 100 times and superimpose power curves
for (iter in 1:10){

# Draw random values in a uniform -1.96 - 1.96 
rand=matrix( runif(n = nvar * nvar, min = -1.96, max = 1.96), nrow = nvar, ncol = nvar)
rand[upper.tri(rand)]=0

randE=matrix( runif(n = nvar * nvar, min = -1.96, max = 1.96), nrow = nvar, ncol = nvar)
randE[upper.tri(randE)]=0

aPathsS= aPaths + rand * aPathsSE
ePathsS= ePaths + randE * ePathsSE

pw1=estimatePowerMultivariateConservative(R2 = R2s[iii], MAF = 0.2, aPaths = aPathsS, ePaths = ePathsS, nvar = 7, Ns = c(1000,2000,3000,4000,5000,6000,7000,8000,9000,10000,15000,20000,25000,30000,35000,40000,45000,50000,55000, 60000), alpha = 5e-8, method = "weights")
lines(c(0,Ns), pw1[1,], col=colM[iii])
}
    pw01=estimatePowerMultivariateConservative(R2 = R2s[iii], MAF = 0.2, aPaths = aPaths, ePaths = ePaths, nvar = 7, Ns = Ns, alpha = 5e-8, method = "weights")
 points(c(0,Ns), pw01[1,], pch=20, col="black")
}

abline(h=0.8, lty=2, lwd=2)
grid()
par(xpd=T)
legend(x = -3000/1000, y = -0.3, legend = c("R2=1%","R2=0.75%", "R2=0.5%", "R2=0.25%", "R2=0.1%", "R2=0.075%", "R2=0.05%"), col=colM, pch=19, cex=1, horiz = T, box.col = "white")
#dev.off()

```

<br><br>
